---
title: Session2a - Data wrangingling with tidyverse
source: Rmd
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Usage and Adaptation of Data Carpentry Materials:
Most material found in this document has been adapted from [Data Carpentry][https://datacarpentry.org/r-socialsci/] materials, under the [creative commons attribution license][https://creativecommons.org/licenses/by/4.0/]. Minor amendments have been made to allow for compatability in order.

### Exercise 0

Go ahead and load in tidyverse as usual.

```{r Tidyverse and Here Loading}
library(tidyverse)
```

-------------

For this workshop, we will need to import a tidy data set. We will continue to use the data set from this morning.

```{r DataImport}
interviews <- read_csv("https://raw.githubusercontent.com/datacarpentry/r-socialsci/main/episodes/data/SAFI_clean.csv")
```

### Exercise 1

Using pipes, subset the `interviews` data to include interviews where respondents were members of an irrigation association (`memb_assoc`) and retain only the columns `affect_conflicts`, `liv_count`, and `no_meals`. Save this in an object called 'interviews_restricted'
```{r Subsetting Task}
interviews %>%
    filter(memb_assoc == "yes") %>%
    select(affect_conflicts, liv_count, no_meals)
```

-------------

### Exercise 2

Create a new dataframe from the `interviews` data that meets the following criteria: contains only the `village` column and a new column called `total_meals` containing a value that is equal to the total number of meals served in the household per day on average (`no_membrs` times `no_meals`). Only the rows where `total_meals` is greater than 20 should be shown in the final dataframe. This should be saved as 'interview_total_meals'

**Hint**: think about how the commands should be ordered to produce this data frame.

```{r New Variable Task}
interviews_total_meals <- interviews %>%
    mutate(total_meals = no_membrs * no_meals) %>%
    filter(total_meals > 20) %>%
    select(village, total_meals)
```

-------------

### Exercise 3

How many households in the survey have an average of two meals per day? Three meals per day? Are there any other numbers of meals represented? 

```{r Counting Task}
interviews %>%
   count(no_meals)
```

-------------

### Exercise 4

Use `group_by()` and `summarise()` to find the mean, min, and max number of household members for each village. Also add the number of observations (hint: see `?n`).

```{r Summary Task}
interviews %>%
  group_by(village) %>%
  summarize(
      mean_no_membrs = mean(no_membrs),
      min_no_membrs = min(no_membrs),
      max_no_membrs = max(no_membrs),
      n = n()
```

-------------

### Exercise 5

Create a new dataframe (named `interviews_months_lack_food`) that has one column for each month and records `TRUE` or `FALSE` for whether each interview respondent was lacking food in that month.

```{r Pivoting Task}
interviews_months_lack_food <- interviews %>%
  separate_longer_delim(months_lack_food, delim = ";") %>%
  mutate(months_lack_food_logical  = TRUE) %>%
  pivot_wider(names_from = months_lack_food,
              values_from = months_lack_food_logical,
              values_fill = list(months_lack_food_logical = FALSE))
```