---
title: Session1b - Working with data in R
source: Rmd
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Usage and Adaptation of Data Carpentry Materials:
Most material found in this document has been adapted from [Data Carpentry][https://datacarpentry.org/r-socialsci/] materials, under the [creative commons attribution license][https://creativecommons.org/licenses/by/4.0/]. Minor amendments have been made to allow for compatability in order.

### Exercise 0

Having installed tidyverse and here packages earlier, we must still load in the packages for our system to use it. Go ahead and do this.

```{r Tidyverse and Here Loading}

```

-------------

For this workshop, we will need to import a tidy data set. We will choose one from the data carpentry.

```{r DataImport}
interviews <- read_csv("https://raw.githubusercontent.com/datacarpentry/r-socialsci/main/episodes/data/SAFI_clean.csv")
```

### Exercise 1
Now you have the data imported, try some of the data summary functions that we discussed on this data. Once you've done that write a short paragraph detailing how many entries there are, how many vairables and what their types are.

```{r Here Install and Loading}
# <INSERT YOUR ANSWER ON LINES BELOW>

```

-------------

### Exercise 2
Some of the variables in the dataframe are encoded as strings, when in fact they are categories. Go ahead and change one of the variables that is like this into a factor.

```{r Factor Change}
# <INSERT YOUR ANSWER ON LINES BELOW>

```

-------------

### Exercise 3
In the tutorial we saw we could dissect the variable 'interview_date' into each of the year, month and date components. Go ahead and replicate this saving each variable under a different name. 

```{r Interview Date Dissection}
# <INSERT YOUR ANSWER ON LINES BELOW>

```